SIMD\slash SIMT-vectorized Sparse Chemical Kinetic Jacobian\slash Thermo-chemical Source-Term Evaluation

### 2018-03-22 paper.tex
rewrite lit review to make it less similar to past works

Recognizing the need to accelerate Jacobian evaluation and factorization, a number of recent works have been published on analytical chemical kinetic Jacobian evaluation; although as will be discussed at the end of this section, this work offers several key improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first packages developed providing this functionality, but has several limitations including: incompatibilities with modern reaction types---i.e., pressure dependent Arrhenius (or P-Log) and Chebyshev reactions---and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although the chosen isothermal assumption is not typical in most combustion simulations, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expense of strict correctness), and enable tabulation of temperature dependent properties; a \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) using a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition, since the Jacobian was based on an over-constrained system, the effect on strict conservation of mass\slash energy was not studied.

A number of recent works have investigated the use of high-performance SIMT-devices to accelerate reactive-flow and chemical kinetics simulations.
Spafford et al.~\cite{Spafford:2010aa} investigated an implementation of an explicit direct numerical simulation code for compressible turbulent combustion on a Tesla C1060 GPU, and found an order of magnitude speedup in evaluation of species production rates as compared to a serial CPU implementation on an AMD-Opteron processor.
Shi et al.~\cite{Shi:2011aa} developed a code to evaluate species production rates and factorize the chemical kinetic Jacobian on a Tesla C2050 GPU, and found an order of magnitude speedup during integration of large chemical models when these operations were offloaded to the GPU as compared to the standard \texttt{CHEMKIN}~\cite{kee1989chemkin} and \texttt{LAPACK}~\cite{Anderson:1999aa} libraries on a quad-core Intel i7 930 CPU; it was not clear how\slash if the CPU code was parallelized.
Niemeyer et al.~\cite{Niemeyer:2011aa} implemented an explicit fourth-order Runge--Kutta integrator for non-stiff chemical kinetic integration on a Tesla C2075 GPU, and found a speedup of nearly two orders of magnitude when compared with a sequential CPU-code on an Intel Xeon X5650 CPU.
Shi et al.~\cite{Shi:2012aa} later extended their work to develop a stabilized-explicit solver on a Tesla C2050, paired with a traditional implicit integrator on quad-core Intel i7 930 which handled the stiffest computational cells; a \SIrange{11}{46}{$\times$} speedup was achieved over the base implicit CPU integrator during simulation of a three-dimensional premixed diesel engine simulation.
Le et al.~\cite{Le2013596} implemented two high-order shock-capturing reactive-flow codes on a Tesla C2070, and found a \numrange{30}{50}$\times$ speedup over a sequential CPU version of the same code on a Intel Xeon X5650.
Stone and Davis~\cite{Stone:2013aa} implemented the implicit VODE~\cite{Brown:1989vl} integrator on a Fermi M2050 GPU, achieving an order of magnitude speedup over the baseline CPU version on a single core of an AMD Opteron 6134 Magny-Cours.
Niemeyer and Sung~\cite{Niemeyer:2014aa} later developed a stabilized explicit second-order Runge--Kutta--Chebyshev algorithm on a Tesla C2075, demonstrating an order of magnitude speedup for a CPU implementation of VODE on a six-core Intel X5650 for moderately stiff chemical kinetics.
Sewerin and Rigopoulos~\cite{Sewerin20151375} implemented a three-stage\slash fifth-order implicit Runge--Kutta method~\cite{wanner1991solving} on both a high-end (NVIDIA Quadro 6000) and consumer-grade (NVIDIA Quadro 600) GPUs, as well as a standard CPU (two-core, four-thread Intel i5-520M) and a scientific workstation (eight-core, 16-thread Intel Xeon E5-2687W); the high-end GPU was at best \SI{1.8}{$\times$} slower than the workstation CPU (16 threads), while the consumer level GPU was at best \SI{5.5}{$\times$} slower than the corresponding standard CPU (four threads).
Yonkee and Sutherland~\cite{Yonkee2016} implemented GPU-accelerated evaluations of thermodynamic parameters, multicomponent transport properties and species production rates for solution of a partial differential equations (PDEs) on an Intel Xeon E5-2680 CPU and a Tesla K20 GPU.
In evaluation of the thermo-chemical properties, speedups between \SIrange{8}{13}{$\times$} were found (over a serial version of the same code) while running on 16 CPU cores, and \SIrange{20}{40}{$\times$} on the GPU; speedups of \textasciitilde\SI{9}{$\times$} and \textasciitilde\SI{25}{$\times$} were found in solution of a partially premixed methanol flame PDE on 16 CPU cores and the GPU, respectively.
Curtis et al.~\cite{CurtisGPU:2017} implemented a fifth-order implicit Runge--Kutta method~\cite{wanner1991solving}, as well as two fourth-order exponential integration techniques~\cite{Hochbruck:1998,Hockbruck:2009} paired with an analytical Jacobian code~\cite{Niemeyer:2016aa} on a Tesla C2075; the implicit Runge--Kutta method was roughly equivalent to a standard implicit integrator~\cite{Hindmarsh:2005} running on 12--38 Intel Xeon E5-4640 v2 CPU cores for two relatively small chemical models with an integration time step of \SI{e-6}{$\sec$}.

>>>>

Recognizing the need to accelerate Jacobian evaluation and factorization, a number of recent works have been published on analytical chemical kinetic Jacobian evaluation; although as will be discussed at the end of this section, this work offers several key improvements over past efforts.
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first codes developed providing analytical Jacobian evaluation, but has several limitations including: incompatibilities with modern reaction types---i.e., pressure dependent Arrhenius (or P-Log) and Chebyshev reactions---and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}. 
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although the chosen isothermal assumption is not typical in most combustion simulations, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expense of strict correctness), and enable tabulation of temperature dependent properties; an \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) using a Tesla C2075 GPU.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition, since the Jacobian was based on an over-constrained system, the effect on strict conservation of mass\slash energy was not studied.

A number of recent works have investigated the use of high-performance SIMT-devices to accelerate reactive-flow and chemical kinetics simulations.
Spafford et al.~\cite{Spafford:2010aa} coupled GPU-based chemical source-term evaluation with an explicit direct numerical simulation code, achieving an order of magnitude speedup on a Tesla C1060 compared to a serial implementation on a AMD-Opteron processor.
Shi et al.~\cite{Shi:2011aa} paired GPU-based chemical source-term and chemical kinetic Jacobian factorization with two implicit solvers, achieving an order of magnitude speedup for homogeneous reactor simulations of large chemical models on a Tesla 2050 GPU, as compared to a serial Intel i7 930 CPU implementation.
Niemeyer et al.~\cite{Niemeyer:2011aa} implemented an explicit solver for non-stiff chemistry on a Tesla C2075 GPU, achieving a speedup of nearly two orders of magnitude over a sequential code on an Intel Xeon X5650 CPU.
Shi et al.~\cite{Shi:2012aa} proposed a strategy for chemical-kinetic integration in three-dimensional reactive-flow simulations, a traditional CPU-based implicit integrator handled the stiffest computational cells while a stabilized-explicit solver on a Tesla C2050 GPU solved the less stiff cells; this hybrid solution technique was \SIrange{11}{46}{$\times$} faster than the implicit CPU solver alone for simulation of a premixed diesel engine.
Le et al.~\cite{Le2013596} found a \SIrange{30}{50}{$\times$} speedup for a GPU-based shock-capturing reactive-flow code on a Tesla C2070 as compared to a sequential version of the same on a Intel Xeon X5650 CPU.
Stone and Davis~\cite{Stone:2013aa} investigated a GPU-version of a common implicit integrator (VODE~\cite{Brown:1989vl}), finding an order of magnitude speedup over a serial CPU implementation.
Niemeyer and Sung~\cite{Niemeyer:2014aa} developed a GPU-based stabilized explicit integrator for use with moderately-stiff chemical kinetics; an order of magnitude was achieved on a Tesla C2075 GPU over a multithreaded VODE solver on a six-core Intel X5650 CPU.
Sewerin and Rigopoulos~\cite{Sewerin20151375} studied an fifth-order implicit Runge-Kutta solver on both consumer-grade and high-end GPUs\slash CPUs; the high-end GPU was at best \SI{1.8}{$\times$} slower than the corresponding CPU.
Yonkee and Sutherland~\cite{Yonkee2016} implemented GPU-accelerated evaluations of thermodynamic parameters, multicomponent transport properties and species production rates for solution of a partial differential equations (PDEs) on an Intel Xeon E5-2680 CPU and a Tesla K20 GPU.
In evaluation of the thermo-chemical properties, speedups between \SIrange{8}{13}{$\times$} were found (over a serial version of the same code) while running on 16 CPU cores, and \SIrange{20}{40}{$\times$} on the GPU; speedups of \textasciitilde\SI{9}{$\times$} and \textasciitilde\SI{25}{$\times$} were found in solution of a partially premixed methanol flame PDE on 16 CPU cores and the GPU, respectively.
Curtis et al.~\cite{CurtisGPU:2017} implemented a fifth-order implicit Runge--Kutta method~\cite{wanner1991solving}, as well as two fourth-order exponential integration techniques~\cite{Hochbruck:1998,Hockbruck:2009} paired with an analytical Jacobian code~\cite{Niemeyer:2016aa} on a Tesla C2075; the implicit Runge--Kutta method was roughly equivalent to a standard implicit integrator~\cite{Hindmarsh:2005} running on 12--38 Intel Xeon E5-4640 v2 CPU cores for two relatively small chemical models with an integration time step of \SI{e-6}{$\sec$}.
